---
title: "NAS_DNAm_Regions_Pathways"
output: html_document
---

This document evaluates CpG overlap and direction of change, generates
DMR-highlighted Manhattan plots, and performs Hallmark pathway enrichment. It
produces Fig. 1c, Fig. 1d, Fig. 2, and Fig. 5.

Run after `NAS_DNAm_Analysis.Rmd` and `limma_corr.R`.

Significant regions are read from Supplementary Data S8–S11 of the article.
Download the Supplementary Data workbook and set `supp_file` below.

# Setup

```{r message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(data.table)
  library(gtools)
  library(ggrepel)
  library(readxl)
})

path_data <- "../Data"
path_out  <- "../Output"

supp_file <- file.path(path_data, "NAS_GCR_DNAm_Supp_Data.xlsx")

source("helper.R")

anno_df <- readRDS(file.path(path_data, "anno_df.RDS"))

exp_tags   <- c(CRII = "CRII", Neutrons = "ntr", SSN = "SSN", IMF = "IMF")
supp_sheet <- c(CRII = "DataS8", Neutrons = "DataS9", SSN = "DataS10", IMF = "DataS11")

ewas <- map(exp_tags, ~ readRDS(file.path(path_out,
              sprintf("EWAS_%s_mov30_results_limma.rds", .x))) %>%
              left_join(anno_df, by = "CpG"))

sig <- map(ewas, ~ filter(.x, q_bh < 0.05))
map_int(sig, nrow)
```

# CpG overlap across indicators

```{r}
sig_sets <- map(sig, "CpG")

overlap_tab <- imap_dfr(sig_sets, function(cpgs, nm) {
  others <- unlist(sig_sets[setdiff(names(sig_sets), nm)], use.names = FALSE)
  tibble(exposure = nm, n_sig = length(cpgs),
         n_unique = sum(!cpgs %in% others),
         pct_unique = round(100 * mean(!cpgs %in% others), 1))
})
write.csv(overlap_tab, file.path(path_out, "Fig1c_cpg_uniqueness.csv"), row.names = FALSE)

overlap_tab
```

```{r}
suppressPackageStartupMessages(library(ggvenn))

p <- ggvenn(sig_sets, fill_color = c("#9b5de5", "#f15bb5", "#00bbf9", "#fee440"),
            stroke_size = 0.5, set_name_size = 4, show_percentage = FALSE) +
  plot.format +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        panel.border = element_blank())

ggsave(p, filename = file.path(path_out, "Fig1c_cpg_overlap.svg"), width = 6, height = 5)
```

# direction of methylation change

```{r}
direction_tab <- imap_dfr(sig, ~ tibble(
  exposure   = .y,
  n_sig      = nrow(.x),
  n_higher   = sum(.x$beta > 0),
  n_lower    = sum(.x$beta < 0),
  pct_higher = round(100 * mean(.x$beta > 0), 1)
))
write.csv(direction_tab, file.path(path_out, "Fig1d_direction.csv"), row.names = FALSE)

direction_tab
```

```{r}
p <- direction_tab %>%
  select(exposure, n_higher, n_lower) %>%
  pivot_longer(-exposure, names_to = "direction", values_to = "n") %>%
  mutate(direction = recode(direction, n_higher = "Higher methylation",
                                       n_lower  = "Lower methylation")) %>%
  ggplot(aes(exposure, n, fill = direction)) +
  geom_col(position = "fill", width = 0.7) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Higher methylation" = "#C1121F",
                               "Lower methylation"  = "#2C7BB6")) +
  labs(x = NULL, y = "Proportion of significant CpGs", fill = NULL) +
  plot.format

ggsave(p, filename = file.path(path_out, "Fig1d_direction.svg"), width = 5, height = 3.5)
```

# DMR-highlighted Manhattan

Each worksheet contains a title row above the column header (`skip = 1`). Region
p values and FDR values are imported as text and converted to numeric values.

```{r}
read_supp_dmr <- function(sheet) {
  read_excel(supp_file, sheet = sheet, skip = 1) %>%
    mutate(across(c(start, end, nprobe), as.numeric),
           across(c(p, fdr), as.numeric),
           chr = gsub("^chr", "", chr)) %>%
    filter(!is.na(start), !is.na(end))
}

dmr <- map(supp_sheet, read_supp_dmr)

map_int(dmr, nrow)
```

```{r}
manhattan_dmr <- function(ewas_i, dmr_i, label) {
  gene_labels <- ewas_i %>%
    mutate(gene_clean = if_else(
      is.na(gene) | gene == "", CpG,
      str_split(gene, ";") |> map_chr(~ unique(.x[.x != ""])[1])
    )) %>%
    select(CpG, gene_clean)

  dat <- ewas_i %>%
    filter(!is.na(chr), !is.na(pos), !is.na(p_bacon)) %>%
    transmute(probe = CpG, p = p_bacon, chr = gsub("^chr", "", chr),
              start = pos, end = pos) %>%
    left_join(gene_labels, by = c("probe" = "CpG")) %>%
    mutate(chr = factor(chr, levels = mixedsort(unique(chr)))) %>%
    arrange(chr, start) %>%
    group_by(chr) %>% mutate(chr_len = max(start, na.rm = TRUE)) %>% ungroup() %>%
    mutate(
      bp_cum      = start + cumsum(as.numeric(chr_len)) - as.numeric(chr_len),
      logp        = -log10(p),
      color_group = as.numeric(chr) %% 2
    ) %>%
    as.data.table()

  reg <- as.data.table(dmr_i)
  # foverlaps requires matching data types for the join columns
  reg[, chr := factor(as.character(chr), levels = levels(dat$chr))]
  setkey(dat, chr, start, end)
  setkey(reg, chr, start, end)

  hits <- foverlaps(dat, reg, nomatch = 0L, which = TRUE)$xid
  dat[, is_sig := FALSE]
  if (length(hits)) dat[hits, is_sig := TRUE]

  top <- dat[is_sig == TRUE][order(p)]
  top <- top[seq_len(min(10, nrow(top)))]
  dat[, is_top := probe %in% top$probe]

  axis_df <- dat[, .(center = mean(range(bp_cum))), by = chr] %>%
    mutate(chr_num = suppressWarnings(as.numeric(as.character(chr))),
           lab     = if_else(!is.na(chr_num) & chr_num > 16 & chr_num %% 2 == 0,
                             "", as.character(chr_num)))

  pd <- as_tibble(dat)
  p <- ggplot() +
    geom_point(data = filter(pd, !is_sig),
               aes(bp_cum, logp, colour = factor(color_group)), size = 0.6, alpha = 0.6) +
    geom_point(data = filter(pd, is_sig & !is_top),
               aes(bp_cum, logp), colour = "#E6521F", size = 0.8, alpha = 0.8) +
    geom_point(data = filter(pd, is_top),
               aes(bp_cum, logp), colour = "#EA2F14", shape = 17, size = 2) +
    geom_text_repel(data = as_tibble(top),
                    aes(bp_cum, logp, label = gene_clean), size = 3, max.overlaps = 1e3) +
    geom_hline(yintercept = -log10(1e-6), linetype = "dashed") +
    scale_colour_manual(values = c("0" = "#2C7BB6", "1" = "#A6CEE3")) +
    scale_x_continuous(breaks = axis_df$center, labels = axis_df$lab, expand = c(0.01, 0)) +
    labs(x = "Chromosome", y = expression(-log[10](p)), title = label) +
    plot.format +
    theme(legend.position = "none")

  ggsave(p, filename = file.path(path_out, sprintf("manhattan_%s.png", tolower(label))),
         width = 8, height = 4)
}

iwalk(exp_tags, ~ manhattan_dmr(ewas[[.y]], dmr[[.y]], .y))
```

# Hallmark pathway enrichment

`methylRRA` uses the bacon- and BH-corrected FDR for every tested CpG and
accounts for CpG-to-gene mapping multiplicity and probe-count bias. The analysis
is restricted to 14 prespecified genome integrity-related Hallmark sets.

```{r}
suppressPackageStartupMessages({
  library(methylGSA)
  library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
})

hallmark <- readRDS(url("http://bioinf.wehi.edu.au/MSigDB/v7.1/Hs.h.all.v7.1.entrez.rds"))

hall_subset <- c(
  "HALLMARK_DNA_REPAIR", "HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY", "HALLMARK_P53_PATHWAY",
  "HALLMARK_OXIDATIVE_PHOSPHORYLATION", "HALLMARK_PEROXISOME",
  "HALLMARK_MTORC1_SIGNALING", "HALLMARK_PI3K_AKT_MTOR_SIGNALING",
  "HALLMARK_UNFOLDED_PROTEIN_RESPONSE", "HALLMARK_PROTEIN_SECRETION",
  "HALLMARK_APOPTOSIS", "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
  "HALLMARK_INTERFERON_GAMMA_RESPONSE", "HALLMARK_INTERFERON_ALPHA_RESPONSE",
  "HALLMARK_INFLAMMATORY_RESPONSE"
)

run_gsea <- function(ewas_i) {
  pv <- ewas_i$q_bh
  names(pv) <- ewas_i$CpG
  set.seed(1996)
  methylRRA(cpg.pval = pv, method = "GSEA", GS.list = hallmark,
            GS.idtype = "ENTREZID", minsize = 30, maxsize = 300, sig.cut = 0.05)
}

gsea <- map(ewas, run_gsea)
write.csv(bind_rows(gsea, .id = "exposure"),
          file.path(path_out, "Fig5_hallmark_gsea.csv"), row.names = FALSE)
```

```{r}
# Restore standard acronym capitalization after applying toTitleCase
tidy_labels <- function(x) {
  x <- gsub("^HALLMARK_", "", x)
  x <- tools::toTitleCase(tolower(gsub("_", " ", x)))
  x <- sub("^Mtorc1", "mTORC1", x)
  x <- sub("^Pi3k Akt Mtor", "PI3K/AKT/mTOR", x)
  x <- sub("^Tnfa", "TNFA", x)
  x <- sub("^Dna", "DNA", x)
  x <- sub("Nfkb$", "NFKB", x)
  x
}

# Figure 5 encodes -log10(FDR) as wedge radius and NES as fill color
iwalk(gsea, function(res, nm) {
  res_circ <- res %>%
    filter(ID %in% hall_subset) %>%
    mutate(ID_clean = tidy_labels(ID))

  p <- circ_hallmark(
    res_circ,
    y_var    = "neglogFDR",
    fill_var = "NES",
    title    = paste("Enriched Hallmark Pathways for", nm),
    sort_by  = "padj"
  )
  ggsave(p, filename = file.path(path_out, sprintf("Fig5_hallmark_%s.svg", tolower(nm))),
         width = 5, height = 5)
})
```
