---
title: "NAS_DNAm_Analysis — preprocessing and epigenetic age"
output: html_document
---

This document performs preprocessing, surrogate-variable estimation, epigenetic
clock analysis, and phenotype-level mixed modeling. 

Region- and pathway-level analyses are in `NAS_DNAm_Regions_Pathways.Rmd`, which
runs after `limma_corr.R`.

Exposures are the 30-day moving average of each indicator, scaled by its
interquartile range.

# Setup

```{r message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(matrixStats)
})

config <- list(
  path_data = "../Data",
  path_out  = "../Output",
  path_code = ".",
  arraytype = "450K",
  norm_method = "BMIQ",
  n_cores = 10,
  seed = 1996,
  n_sv = 8,
  top_var_probes = 40000
)

set.seed(config$seed)

path_data <- config$path_data
path_out  <- config$path_out
path_code <- config$path_code

source(file.path(path_code, "helper.R"))

beta_raw <- readRDS(file.path(path_data, "beta2_impu_cleaning_champnorm.RDS"))
pheno    <- readRDS(file.path(path_data, "pheno_methy.RDS"))

stopifnot(is.data.frame(pheno), "samplename_450k" %in% names(pheno))
```

# Data processing

## Normalization and M-values

Beta values are normalized using BMIQ in ChAMP and logit-transformed to M-values.

```{r}
suppressPackageStartupMessages(library(ChAMP))

beta_norm <- champ.norm(
  beta      = beta_raw,
  arraytype = config$arraytype,
  method    = config$norm_method,
  plotBMIQ  = TRUE,
  cores     = config$n_cores
)
saveRDS(beta_norm, file.path(path_data, "beta_norm_champ.RDS"))

epsilon   <- 1e-6
beta_clip <- pmin(pmax(beta_norm, epsilon), 1 - epsilon)
dnam_mval <- log2(beta_clip / (1 - beta_clip))

common_samples <- intersect(pheno$samplename_450k, colnames(dnam_mval))
pheno     <- pheno[match(common_samples, pheno$samplename_450k), , drop = FALSE]
dnam_mval <- dnam_mval[, match(common_samples, colnames(dnam_mval)), drop = FALSE]
stopifnot(identical(pheno$samplename_450k, colnames(dnam_mval)))

message(nrow(dnam_mval), " probes × ", ncol(dnam_mval), " samples, ",
        length(unique(pheno$ID)), " participants")
```

## Surrogate variables (SmartSVA)

Eight SVs are estimated with the design matrix below; `num.sv` is reported for
comparison. 

```{r}
suppressPackageStartupMessages({
  library(sva)
  library(SmartSVA)
})

pheno <- pheno %>% mutate(batch = sub("_.*$", "", samplename_450k))
rownames(pheno) <- pheno$samplename_450k

idx_top <- order(rowVars(dnam_mval), decreasing = TRUE)[seq_len(config$top_var_probes)]
Msub    <- dnam_mval[idx_top, , drop = FALSE]

mod <- model.matrix(
  ~ CRII.mov365 + ntr.mov365 + PLB.mov365 + SSN.mov365 + IMF.mov365 + Kp_index.mov365 +
    AGE + SMK + packyrs + bmi + educmax + temp_c + VISIB + year + season +
    CD8T_est450k + CD4T_est450k + NK_est450k + Bcell_est450k + Mono_est450k + Gran_est450k,
  data = pheno
)
mod0 <- model.matrix(
  ~ AGE + SMK + packyrs + bmi + educmax + temp_c + VISIB + year + season +
    CD8T_est450k + CD4T_est450k + NK_est450k + Bcell_est450k + Mono_est450k + Gran_est450k,
  data = pheno
)

message("Suggested n.sv (be, B=25): ", num.sv(Msub, mod, method = "be", B = 25))

svobj <- smartsva.cpp(dat = dnam_mval, mod = mod, mod0 = mod0, n.sv = config$n_sv)
saveRDS(svobj, file.path(path_data, sprintf("sva_sv%d.RDS", config$n_sv)))

sv_cols <- paste0("sv", seq_len(config$n_sv))
pheno <- pheno %>%
  select(-any_of(paste0("sv", 1:20))) %>%
  bind_cols(as.data.frame(svobj$sv))
colnames(pheno)[(ncol(pheno) - config$n_sv + 1):ncol(pheno)] <- sv_cols

saveRDS(dnam_mval, file.path(path_data, "dnam_mval.RDS"))
```

## Annotation

```{r}
suppressPackageStartupMessages({
  library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
  library(minfi)
})

anno <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
anno_df <- tibble(
  CpG    = rownames(anno),
  chr    = as.character(anno$chr),
  pos    = as.integer(anno$pos),
  gene   = as.character(anno$UCSC_RefGene_Name),
  island = as.character(anno$Relation_to_Island)
)
saveRDS(anno_df, file.path(path_data, "anno_df.RDS"))
```

## Epigenetic clocks

Clock values are obtained from the Clock Foundation calculator 
(https://dnamage.clockfoundation.org). EAA is calculated as the
residual from regressing each clock on chronological age, cell fractions, and
the eight SVs.

```{r}
DNAmAge <- readr::read_csv(
  file.path(path_data, "DNAmAgeCalcProject_19645_Results.csv"),
  show_col_types = FALSE
)

old_names  <- c("DNAmAge", "DNAmAgeHannum", "DNAmPhenoAge", "DNAmGrimAge2BasedOnRealAge")
dnam_clock <- c("HorvathAge", "HannumAge", "PhenoAge", "GrimAge")
names(DNAmAge)[match(old_names, names(DNAmAge))] <- dnam_clock

DNAmAge <- DNAmAge %>% mutate(samplename_450k = gsub("S", "", SID))

adj_terms <- c("Age", "CD8T", "CD4T", "NK", "Bcell", "Mono", "Gran", sv_cols)
for (clk in dnam_clock) {
  fml <- as.formula(paste(clk, "~", paste(adj_terms, collapse = "+")))
  DNAmAge[[paste0(clk, "_acc")]] <- residuals(lm(fml, data = DNAmAge))
}
dnam_acc <- paste0(dnam_clock, "_acc")

pheno <- pheno %>%
  left_join(
    DNAmAge %>% select(samplename_450k,
                       any_of(c(dnam_clock, dnam_acc,
                                "CD8T", "CD4T", "NK", "Bcell", "Mono", "Gran"))),
    by = "samplename_450k"
  )
```

## Exposure scaling

`run_lmer_loop_iqr()` scales internally; the `_IQR` columns written here are for
the EWAS design in `limma_corr.R`.

```{r}
indicators <- c("CRII", "ntr", "SSN", "IMF", "Kp_index")
exposures  <- paste0(indicators, ".mov30")

pheno_df <- pheno %>%
  mutate(across(all_of(exposures), ~ . / IQR(., na.rm = TRUE), .names = "{.col}_IQR"))

saveRDS(pheno_df, file.path(path_data, "pheno_methy_Aug2025.RDS"))
```

# Phenotype-based analyses

## Table 1

```{r}
suppressPackageStartupMessages(library(tableone))

baseline <- pheno_df %>%
  arrange(ID, date) %>% group_by(ID) %>% slice(1) %>% ungroup()

vars     <- c("AGE","bmi","packyrs","educmax","SMK","diabete","Statin_flag","med","WBC","NEUT","LYMPH")
cat_vars <- c("SMK","diabete","Statin_flag","med")
print(CreateTableOne(vars = vars, data = baseline, factorVars = cat_vars),
      nonnormal = "packyrs")
```

## Exposure IQRs by averaging window

```{r}
sapply(c(daily = "", `7-day` = ".mov7", `30-day` = ".mov30"), function(sfx) {
  sapply(indicators, function(ind) IQR(pheno_df[[paste0(ind, sfx)]], na.rm = TRUE))
}) %>% round(2)
```

## Mixed models

Three sequential covariate sets. Model 3 adds temperature, visibility and season.

```{r}
outcomes   <- dnam_acc
cell_terms <- sprintf("scale(%s)", c("CD8T","CD4T","NK","Bcell","Mono","Gran"))

cov_mod1 <- c("AGE", "year", cell_terms, sv_cols)
cov_mod2 <- c(cov_mod1, "bmi", "as.factor(SMK)", "packyrs", "educmax",
              "Statin_flag", "med", "diabete")
cov_mod3 <- c(cov_mod2, "temp_c.mov30", "VISIB.mov30", "season")

forest_df <- bind_rows(
  run_lmer_loop_iqr(pheno_df, outcomes, exposures, cov_mod1) %>% mutate(model = "mod1"),
  run_lmer_loop_iqr(pheno_df, outcomes, exposures, cov_mod2) %>% mutate(model = "mod2"),
  run_lmer_loop_iqr(pheno_df, outcomes, exposures, cov_mod3) %>% mutate(model = "mod3")
)
write.csv(forest_df, file.path(path_out, "forest_mov30_age.csv"), row.names = FALSE)

forest_df %>% filter(model == "mod3", outcome == "HorvathAge_acc")
```

## Fig. 1b

```{r}
suppressPackageStartupMessages(library(grid))

plot_df <- forest_df %>%
  mutate(
    exposure = gsub("\\.mov\\d+$", "", exposure),
    exposure = factor(exposure, levels = rev(unique(exposure))),
    model    = factor(model, levels = rev(c("mod1","mod2","mod3"))),
    outcome  = factor(outcome, levels = outcomes)
  )

p <- ggplot(plot_df, aes(exposure, estimate, ymin = conf.low, ymax = conf.high, color = model)) +
  geom_errorbar(position = position_dodge(width = 0.6), width = 0.2, linewidth = 0.6) +
  geom_point(position = position_dodge(width = 0.6), size = 2.2) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.6) +
  coord_flip() +
  facet_wrap(~ outcome, ncol = length(outcomes)) +
  scale_color_manual(values = c(mod1 = "#78B7D0", mod2 = "#227B94", mod3 = "#16325B")) +
  labs(x = NULL, y = "EAA (95% CI)", color = "Model") +
  plot.format +
  theme(panel.spacing = unit(1, "lines"), legend.position = "none")

ggsave(p, filename = file.path(path_out, "forest_mov30_age.svg"), width = 7.5, height = 2.75)
```
