---
title: "NAS_DNAm_Analysis"
output: html_document
---

# Setup

```{r message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(matrixStats)
})

config <- list(
  path_data = "../Data",
  path_out  = "../Output",
  path_code = "../Scripts",
  arraytype = "450K",
  norm_method = "BMIQ",
  n_cores = 10,
  seed = 1996,
  n_sv = 8,
  top_var_probes = 40000
)

set.seed(config$seed)

path_data <- config$path_data
path_out  <- config$path_out
path_code <- config$path_code

source(file.path(path_code, "helper.R"))

# Inputs
beta_raw <- readRDS(file.path(path_data, "beta2_impu_cleaning(original version 2_probe cleaning_imputation).Rds"))
pheno    <- readRDS(file.path(path_data, "pheno_methy.RDS"))

stopifnot(is.matrix(beta_raw) || is.data.frame(beta_raw))
stopifnot(is.data.frame(pheno))
stopifnot("samplename_450k" %in% names(pheno))
```

# Data processing

## Exposure scaling (IQR units)

```{r}
expo_names <- c(
  "CRII.mov365","ntr.mov365","PLB.mov365",
  "SSN.mov365","IMF.mov365","Kp_index.mov365"
)

for (e in expo_names) {
  q75 <- quantile(pheno[[e]], 0.75, na.rm = TRUE)
  q25 <- quantile(pheno[[e]], 0.25, na.rm = TRUE)
  pheno[[paste0(e, "_IQR")]] <- (pheno[[e]] - q25) / (q75 - q25)
}
```

## DNAm normalization and M-value conversion

```{r}
suppressPackageStartupMessages(library(ChAMP))

beta_norm <- champ.norm(
  beta      = beta_raw,
  arraytype = config$arraytype,
  method    = config$norm_method,
  plotBMIQ  = TRUE,
  cores     = config$n_cores
)

saveRDS(beta_norm, file.path(path_data, "beta_norm_champ.RDS"))

# Convert beta (0–1) to M-values with clipping
epsilon <- 1e-6
beta_clip <- pmin(pmax(beta_norm, epsilon), 1 - epsilon)
dnam_mval <- log2(beta_clip / (1 - beta_clip))

# Keep only matched samples between phenotype and DNAm matrix
common_samples <- intersect(pheno$samplename_450k, colnames(dnam_mval))
pheno <- pheno[match(common_samples, pheno$samplename_450k), , drop = FALSE]
dnam_mval <- dnam_mval[, match(common_samples, colnames(dnam_mval)), drop = FALSE]

stopifnot(identical(pheno$samplename_450k, colnames(dnam_mval)))

message("M-value matrix: ", nrow(dnam_mval), " probes × ", ncol(dnam_mval), " samples")
message("NA count: ", sum(is.na(dnam_mval)))
message("NaN count: ", sum(is.nan(dnam_mval)))
message("Inf count: ", sum(is.infinite(dnam_mval)))
```

## Surrogate variable estimation (SmartSVA)

```{r}
suppressPackageStartupMessages({
  library(sva)
  library(SmartSVA)
})

pheno <- pheno %>%
  mutate(
    batch = sub("_.*$", "", samplename_450k)
  )
rownames(pheno) <- pheno$samplename_450k

# Subset to most variable probes for SV estimation
idx_top <- order(rowVars(dnam_mval), decreasing = TRUE)[seq_len(config$top_var_probes)]
Msub <- dnam_mval[idx_top, , drop = FALSE]

mod <- model.matrix(
  ~ CRII.mov365 + ntr.mov365 + PLB.mov365 + SSN.mov365 + IMF.mov365 + Kp_index.mov365 +
    AGE + SMK + packyrs + bmi + educmax + temp_c + VISIB + year + season +
    CD8T_est450k + CD4T_est450k + NK_est450k + Bcell_est450k + Mono_est450k + Gran_est450k,
  data = pheno
)

mod0 <- model.matrix(
  ~ AGE + SMK + packyrs + bmi + educmax + temp_c + VISIB + year + season +
    CD8T_est450k + CD4T_est450k + NK_est450k + Bcell_est450k + Mono_est450k + Gran_est450k,
  data = pheno
)

# Optional: estimate n.sv empirically (kept for reference in logs)
n_suggested <- num.sv(Msub, mod, method = "be", B = 25)
message("Suggested n.sv (be, B=25): ", n_suggested)

svobj <- smartsva.cpp(dat = dnam_mval, mod = mod, mod0 = mod0, n.sv = config$n_sv)
saveRDS(svobj, file.path(path_data, sprintf("sva_sv%d.RDS", config$n_sv)))

# Replace any existing sv columns and append new ones
sv_cols <- paste0("sv", seq_len(config$n_sv))
pheno <- pheno %>%
  select(-any_of(paste0("sv", 1:20))) %>%  
  bind_cols(as.data.frame(svobj$sv))
colnames(pheno)[(ncol(pheno) - config$n_sv + 1):ncol(pheno)] <- sv_cols

saveRDS(pheno, file.path(path_data, "pheno_methy_with_sv.RDS"))
saveRDS(dnam_mval, file.path(path_data, "dnam_mval.RDS"))
```

## Annotation

```{r}
suppressPackageStartupMessages({
  library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
  library(minfi)
})

anno <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
anno_df <- tibble(
  CpG    = rownames(anno),
  chr    = as.character(anno$chr),
  pos    = as.integer(anno$pos),
  gene   = as.character(anno$UCSC_RefGene_Name),
  island = as.character(anno$Relation_to_Island)
)

saveRDS(anno_df, file.path(path_data, "anno_df.RDS"))
```

## DNAm clocks (age acceleration residuals)

```{r}
DNAmAge <- readr::read_csv("DNAmAgeCalcProject_19645_Results.csv", show_col_types = FALSE)

old_names  <- c("DNAmAge", "DNAmAgeHannum", "DNAmPhenoAge", "DNAmGrimAge2BasedOnRealAge")
dnam_clock <- c("HorvathAge", "HannumAge", "PhenoAge", "GrimAge")
names(DNAmAge)[match(old_names, names(DNAmAge))] <- dnam_clock

# Harmonize sample ID field from calculator output
DNAmAge <- DNAmAge %>%
  mutate(
    samplename_450k = gsub("S", "", SID)
  )

sv_terms <- paste0("sv", seq_len(config$n_sv))
adj_terms <- c("Age", "CD8T", "CD4T", "NK", "Bcell", "Mono", "Gran", sv_terms)

for (clk in dnam_clock) {
  fml <- as.formula(paste(clk, "~", paste(adj_terms, collapse = "+")))
  DNAmAge[[paste0(clk, "_acc")]] <- residuals(lm(fml, data = DNAmAge))
}

dnam_acc <- paste0(dnam_clock, "_acc")

# Merge clock outputs onto phenotype table (if samplename_450k matches)
pheno <- pheno %>%
  left_join(
    DNAmAge %>% select(samplename_450k, any_of(c(dnam_clock, dnam_acc, "CD8T","CD4T","NK","Bcell","Mono","Gran"))),
    by = "samplename_450k"
  )

# Also compute accelerations directly within pheno using clinical AGE (optional)
adj_terms_pheno <- c("AGE", "CD8T", "CD4T", "NK", "Bcell", "Mono", "Gran", sv_terms)
for (clk in dnam_clock) {
  if (!clk %in% names(pheno)) next
  fml <- as.formula(paste(clk, "~", paste(adj_terms_pheno, collapse = "+")))
  pheno[[paste0(clk, "_acc_pheno")]] <- residuals(lm(fml, data = pheno))
}
```

# Phenotype-based analyses

```{r}
pheno_df <- pheno

vars_to_scale <- c("CRII.mov30", "ntr.mov30", "SSN.mov30", "IMF.mov30", "Kp_index.mov30", "PLB.mov30")
pheno_df <- pheno_df %>%
  mutate(
    across(all_of(vars_to_scale), ~ . / IQR(., na.rm = TRUE), .names = "{.col}_IQR")
  )

saveRDS(pheno_df, file.path(path_data, "pheno_methy_Aug2025.RDS"))
```

## Descriptive statistics

```{r}
suppressPackageStartupMessages(library(tableone))

pheno_df <- pheno_df %>%
  arrange(ID, date) %>%
  group_by(ID) %>%
  mutate(visit_number = row_number()) %>%
  ungroup()

table(pheno_df$visit_number)

vars <- c("AGE","bmi","packyrs","educmax","SMK","diabete","Statin_flag","med","WBC","NEUT","LYMPH")
cat_vars <- c("SMK","diabete","Statin_flag","med")
table1 <- CreateTableOne(vars = vars, data = pheno_df %>% filter(visit_number == 1), factorVars = cat_vars)
print(table1, nonnormal = c("packyrs"))
```

## Regression 

```{r}
## Function to run lmer models in a loop 
library(lmerTest)
library(broom.mixed)
library(dplyr)

run_lmer_loop_iqr <- function(data, outcomes, exposures, covariates, random_effect = "ID") {
  results <- list()
  
  for (x in exposures) {
    iqr_val <- IQR(data[[x]], na.rm = TRUE)
    if (iqr_val == 0 || is.na(iqr_val)) next
    
    scaled_var <- paste0(x, "_scaled")
    data[[scaled_var]] <- data[[x]] / iqr_val
    
    for (y in outcomes) {
      rhs <- paste(c(scaled_var, covariates), collapse = " + ")
      formula_str <- paste0(y, " ~ ", rhs, " + (1 | ", random_effect, ")")
      formula <- as.formula(formula_str)
      
      mod <- tryCatch(
        lmer(formula, data = data, REML = FALSE),
        error = function(e) return(NULL)
      )
      
      if (!is.null(mod)) {
        tidy_mod <- broom.mixed::tidy(mod, effects = "fixed", conf.int = TRUE) %>%
          filter(term == scaled_var) %>%
          mutate(
            outcome = y,
            exposure = x,
            iqr = iqr_val,
            star = case_when(
              p.value < 0.001 ~ "***",
              p.value < 0.01  ~ "**",
              p.value < 0.05  ~ "*",
              p.value < 0.1   ~ ".",
              TRUE            ~ ""
            )
          ) %>%
          dplyr::select(outcome, exposure, iqr, estimate, std.error, conf.low, conf.high, p.value, star)
        
        results[[paste0(y, "_", x)]] <- tidy_mod
      }
    }
  }
  
  final_df <- bind_rows(results)
  return(final_df)
}
```

```{r}
exposures <- c("CRII", "ntr", "SSN", "IMF", "Kp_index")
outcomes  <- dnam_acc

sv_cols <- paste0("sv", seq_len(config$n_sv))

cov_mod3 <- c(
  "AGE","year","bmi","as.factor(SMK)","packyrs","educmax","Statin_flag","med","diabete",
  "temp_c","VISIB","season",
  "scale(CD8T)","scale(CD4T)","scale(NK)","scale(Bcell)","scale(Mono)","scale(Gran)",
  sv_cols
)

cov_mod2 <- c(
  "AGE","year","bmi","as.factor(SMK)","packyrs","educmax","Statin_flag","med","diabete",
  "scale(CD8T)","scale(CD4T)","scale(NK)","scale(Bcell)","scale(Mono)","scale(Gran)",
  sv_cols
)

cov_mod1 <- c(
  "AGE","year",
  "scale(CD8T)","scale(CD4T)","scale(NK)","scale(Bcell)","scale(Mono)","scale(Gran)",
  sv_cols
)

mod3 <- run_lmer_loop_iqr(data = pheno_df, outcomes = outcomes, exposures = exposures, covariates = cov_mod3) %>% mutate(model = "mod3")
mod2 <- run_lmer_loop_iqr(data = pheno_df, outcomes = outcomes, exposures = exposures, covariates = cov_mod2) %>% mutate(model = "mod2")
mod1 <- run_lmer_loop_iqr(data = pheno_df, outcomes = outcomes, exposures = exposures, covariates = cov_mod1) %>% mutate(model = "mod1")

forest_df <- bind_rows(mod1, mod2, mod3)
```

## Forest plot

```{r}
suppressPackageStartupMessages(library(grid))

forest_df <- forest_df %>%
  mutate(
    exposure = gsub("\\.mov\\d+$", "", exposure),
    exposure = factor(exposure, levels = rev(unique(exposure))),
    model    = factor(model, levels = rev(c("mod1","mod2","mod3"))),
    outcome  = factor(outcome, levels = outcomes)
  )

dodge_width <- 0.6
p <- ggplot(forest_df, aes(x = exposure, y = estimate, ymin = conf.low, ymax = conf.high, color = model)) +
  geom_errorbar(position = position_dodge(width = dodge_width), width = 0.2, linewidth = 0.6) +
  geom_point(position = position_dodge(width = dodge_width), size = 2.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", alpha = 0.6) +
  coord_flip() +
  facet_wrap(~ outcome, ncol = length(outcomes)) +
  scale_color_manual(values = c("mod1" = "#78B7D0", "mod2" = "#227B94", "mod3" = "#16325B")) +
  labs(x = NULL, y = "EAA (95% CI)", color = "Model") +
  plot.format +
  theme(panel.spacing = unit(1, "lines"), strip.text = element_text(face = "bold"), legend.position = "none")

ggsave(p, filename = file.path(path_out, "forest_daily_age_kp.svg"), width = 7.5, height = 2.75)
```

# DNAm analyses

## EWAS results post-processing

```{r}
anno_df <- readRDS(file.path(path_data, "anno_df.RDS"))

annotate_ewas <- function(df, anno_df) {
  df %>%
    mutate(se = beta / t) %>%
    left_join(anno_df, by = c("CpG" = "CpG"))
}

# Load EWAS objects (run in background)
ewas_list <- list(
  CRII    = EWAS_CRII_results,
  ntr     = EWAS_ntr_results,
  SSN     = EWAS_SSN_results,
  IMF     = EWAS_IMF_results,
  Kp      = EWAS_Kp_results
)

ewas_list <- imap(ewas_list, ~ annotate_ewas(.x, anno_df))

# Save annotated results
iwalk(ewas_list, ~ write.csv(.x, file.path(path_out, paste0("EWAS_", .y, "_results.csv")), row.names = FALSE))

# Inflation factors (t statistics)
suppressPackageStartupMessages(library(bacon))
iwalk(ewas_list, ~ message(.y, " inflation: ", inflation(bacon(.x$t))))
```

## CpG overlap (Venn) and QQ plots

```{r}
suppressPackageStartupMessages({
  library(ggvenn)
})

venn_data <- list(
  CRII     = ewas_list$CRII %>% filter(q_bh < 0.05) %>% pull(CpG),
  Neutrons = ewas_list$ntr  %>% filter(q_bh < 0.05) %>% pull(CpG)
)

p <- ggvenn(venn_data, fill_color = c("#9b5de5", "#f15bb5"), stroke_size = 0.5, set_name_size = 4) +
  labs(title = "Significant CpGs: CRII vs Neutrons", x = NULL, y = NULL) +
  plot.format +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), panel.border = element_blank())

ggsave(p, filename = file.path(path_out, "venn_CRII_ntr.svg"), width = 6, height = 4)
```

## DMR analysis (comb-p) + Manhattan highlighting DMR hits 

```{r}
suppressPackageStartupMessages({
  library(ENmix)
  library(gtools)
  library(data.table)
  library(ggrepel)
})

dat_imf <- ewas_list$IMF %>%
  transmute(probe = CpG, p = p_bacon, chr = chr, start = pos, end = pos)

# Run combp 
# dmr_imf <- combp(dat_imf, dist.cutoff = 1000, seed = 0.01, nCores = config$n_cores)

dmr_imf <- readr::read_csv("combp-imf/resu_combp.csv", show_col_types = FALSE) %>%
  filter(fdr < 0.05, nprobe > 2) %>%
  mutate(chr = gsub("^chr", "", chr))

gene_labels <- ewas_list$IMF %>%
  mutate(gene_clean = if_else(
    is.na(gene) | gene == "",
    CpG,
    str_split(gene, ";") |> map_chr(~ unique(.x[.x != ""])[1])
  )) %>%
  select(CpG, gene_clean)

dat2 <- dat_imf %>%
  left_join(gene_labels, by = c("probe" = "CpG")) %>%
  mutate(chr = gsub("^chr", "", chr)) %>%
  mutate(chr = factor(chr, levels = mixedsort(unique(chr)))) %>%
  arrange(chr, start) %>%
  group_by(chr) %>%
  mutate(chr_len = max(start, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    bp_cum = start + cumsum(as.numeric(chr_len)) - as.numeric(chr_len),
    logp = -log10(p),
    color_group = as.numeric(chr) %% 2
  ) %>%
  as.data.table()

setDT(dmr_imf)
dmr_imf[, `:=`(start = as.numeric(start), end = as.numeric(end))]
setkey(dat2, chr, start, end)
setkey(dmr_imf, chr, start, end)

hits_idx <- foverlaps(dat2, dmr_imf, nomatch = 0L, which = TRUE)$xid
dat2[, is_sig := FALSE]
dat2[hits_idx, is_sig := TRUE]

top_hits <- dat2[is_sig == TRUE][order(p)][1:min(10, .N)]
dat2[, is_top := probe %in% top_hits$probe]

axis_df <- dat2[, .(center = mean(range(bp_cum))), by = chr] %>%
  mutate(chr_num = suppressWarnings(as.numeric(as.character(chr))),
         label   = if_else(!is.na(chr_num) & chr_num > 16 & chr_num %% 2 == 0, "", as.character(chr_num)))

plot_df <- as_tibble(dat2)

p <- ggplot() +
  geom_point(data = filter(plot_df, !is_sig), aes(bp_cum, logp, colour = factor(color_group)), size = 0.6, alpha = 0.6) +
  geom_point(data = filter(plot_df, is_sig & !is_top), aes(bp_cum, logp), colour = "#E6521F", size = 0.8, alpha = 0.8) +
  geom_point(data = filter(plot_df, is_top), aes(bp_cum, logp), colour = "#EA2F14", shape = 17, size = 2) +
  geom_text_repel(data = as_tibble(top_hits), aes(bp_cum, logp, label = gene_clean), size = 3, max.overlaps = 1e3) +
  geom_hline(yintercept = -log10(1e-6), linetype = "dashed") +
  scale_colour_manual(values = c("0" = "#2C7BB6", "1" = "#A6CEE3")) +
  scale_x_continuous(breaks = axis_df$center, labels = axis_df$label, expand = c(0.01, 0)) +
  labs(x = "Chromosome", y = expression(-log[10](p)), title = "IMF (DMR-highlighted Manhattan)") +
  plot.format +
  theme(legend.position = "none")

ggsave(p, filename = file.path(path_out, "manhattan_imf.png"), width = 8, height = 4)
```

# Global methylation summaries (beta scale)

```{r}
beta_norm <- readRDS(file.path(path_data, "beta_norm_champ.RDS"))
anno_df   <- readRDS(file.path(path_data, "anno_df.RDS"))

anno <- anno_df
rownames(anno) <- anno$CpG
anno <- anno[match(rownames(beta_norm), rownames(anno)), , drop = FALSE]

keep_auto <- !is.na(anno$chr) & anno$chr %in% paste0("chr", 1:22)
beta_auto <- beta_norm[keep_auto, , drop = FALSE]
anno_auto <- anno[keep_auto, , drop = FALSE]

common <- intersect(colnames(beta_auto), rownames(pheno))
beta_auto <- beta_auto[, common, drop = FALSE]
pheno_g <- pheno[common, , drop = FALSE]
rownames(pheno_g) <- pheno_g$samplename_450k

suppressPackageStartupMessages(library(limma))

sv_cols <- paste0("sv", seq_len(config$n_sv))
covar_cols <- c("CD8T","CD4T","NK","Bcell","Mono","Gran", sv_cols)
covar_cols <- intersect(covar_cols, colnames(pheno_g))

beta_adj <- removeBatchEffect(beta_auto, covariates = as.matrix(pheno_g[, covar_cols, drop = FALSE]))

entropy_beta <- function(M) {
  M <- pmax(pmin(M, 1 - 1e-6), 1e-6)
  -(M * log(M) + (1 - M) * log(1 - M)) / log(2)
}

summarize_global <- function(B, anno, block_n = 50000) {
  labs <- c("All","Island","OpenSea","N_Shore","S_Shore","N_Shelf","S_Shelf")
  idxs <- list(
    All      = rep(TRUE, nrow(anno)),
    Island   = anno$island == "Island",
    OpenSea  = anno$island == "OpenSea",
    N_Shore  = anno$island == "N_Shore",
    S_Shore  = anno$island == "S_Shore",
    N_Shelf  = anno$island == "N_Shelf",
    S_Shelf  = anno$island == "S_Shelf"
  )

  out <- list()
  for (nm in labs) {
    sel <- which(idxs[[nm]] & !is.na(idxs[[nm]]))
    if (!length(sel)) next

    nS <- ncol(B)
    sum_beta <- numeric(nS)
    sum_entropy <- numeric(nS)
    sum_sq <- numeric(nS)
    n_nonNA <- numeric(nS)

    for (st in seq(1, length(sel), by = block_n)) {
      ii <- sel[st:min(st + block_n - 1, length(sel))]
      X <- B[ii, , drop = FALSE]
      notNA <- !is.na(X)
      n_nonNA <- n_nonNA + colSums(notNA)

      X0 <- X; X0[!notNA] <- 0
      sum_beta <- sum_beta + colSums(X0)
      sum_sq   <- sum_sq + colSums(X0 * X0)

      E <- entropy_beta(X)
      E[!notNA] <- 0
      sum_entropy <- sum_entropy + colSums(E)
    }

    mml <- sum_beta / n_nonNA
    sdB <- sqrt(pmax(0, (sum_sq / n_nonNA) - mml^2))
    nme <- sum_entropy / n_nonNA

    out[[nm]] <- tibble(sample = colnames(B), context = nm, MML = mml, NME = nme, SD_beta = sdB)
  }

  bind_rows(out)
}

global_df <- summarize_global(beta_adj, anno_auto)

global_wide <- global_df %>%
  pivot_wider(
    id_cols = sample,
    names_from = context,
    values_from = c(MML, NME, SD_beta),
    names_sep = "_"
  )

pheno <- pheno %>%
  left_join(global_wide, by = c("samplename_450k" = "sample"))

saveRDS(pheno, file.path(path_data, "pheno_methy_Aug2025_with_global.RDS"))
```
